---
title: 'Inference Lab 1: the Bootstrap'
author: "Data Science Team"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(message=FALSE,warning = FALSE, comment="", fig.height=4)
```


## Today's Activity

Today's activity on the [bootstrap](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) will 
start with a demo of how to code a bootstrap for one or two variables. 

```{r}
library(dplyr)
library(ggplot2)
movies <- read.csv("https://web.stanford.edu/class/stats101/data/movies.csv", stringsAsFactors = FALSE)
glimpse(movies)
```

Below is the code for bootstrapping a generic statistic `stat` with data `x`, using
B bootstrap samples:

# Bootstrapping a Univariate Statistic

```{r, eval=FALSE}
# x - data points for boot strap sample 
# stat - function that represents statistic wanting to be calculated from the samples 
# number of bootstrap samples 

boot <- function(x, stat, B=1000){
  
  n <- length(x)
  boot_stats <- matrix(nrow=B)
  
  for(i in 1:B){
    indices <- sample(n, replace=TRUE)
    boot_stats[i] <- stat(x[indices])
  }
  
  return(boot_stats)
}
```

For example, if we want to bootstrap the mean with the data `x` and then compute the 2.5th percentile
and the 97.5th percentile of the bootstrap replicates (we will see in class that this is
a 95 percent bootstrap confidence interval):

```{r, eval=FALSE}
x = rnorm(100)
mean.x.boots <- boot(x, mean, 1000)

quantile(mean.x.boots, c(.025, 0.975))


hist(mean.x.boots)

```

**Task** Choose a genre and create a data frame that only contains that data for that genre (but not one with a trivial number of observations). Use this genre throughout.

```{r, eval=FALSE}
sort(table(movies$genre), decreasing = TRUE)
movies_comedy = subset(movies, genre == "Comedy")
```

**Task** Choose a variable and bootstrap the mean. Now calculate a 95% confidence interval using the quantile function on the bootstrap replicates. Also compute the 95% confidence interval using normal approximation: mean +/- 2 SE,
where the SE is computed with the bootstrap.

``` {r}
comedyVotes = movies_comedy$votes 
mean.comedy.votes = boot(comedyVotes, mean, 1000)

quantile(mean.comedy.votes, c(0.025, 0.975))

SE = sd(mean.comedy.votes)
c(mean(movies_comedy$votes) - 2*SE, mean(movies_comedy$votes) + 2*SE)

hist(mean.comedy.votes)

```

**Question** Now bootstrap the median. Which do you expect to be more robust (stable across bootstrap simulations), the mean or the median, and why? 
``` {r}
median.comedy.votes = boot(comedyVotes, median, 1000)

quantile(median.comedy.votes, c(0.025, 0.975))
hist(median.comedy.votes)
```

## Bootstrapping a Bivariate Statistic

The correlation coefficient that we discussed in class is also called Pearson's correlation.
There is a different correlation coefficient called Spearman's correlation, which is more
robust to outlying values (just as the median is more robust than the mean).

```{r, eval=FALSE}

bi_boot <- function(x, y, stat, B, ...){
  
  n <- length(x)
  boot_stats <- matrix(nrow=B)
  
  for(i in 1:B){
    indices <- sample(n, replace=TRUE) 
    boot_stats[i] <- stat(x[indices], y[indices], ...)
  }
  
  return(boot_stats)
  
}

Nsims <- 10000
pearsons <- bi_boot(movies$gross, movies$budget, cor, Nsims)
spearmans <- bi_boot(movies$gross, movies$budget, cor, Nsims, method = "spearman")

# inspect quantiles
pearsons %>% quantile(c(0.025, .1, .5, .9, 0.975)) %>% round(4) 
spearmans %>% quantile(c(0.025, .1, .5, .9, 0.975)) %>% round(4)

# or summary stats
mean(pearsons)
sd(pearsons)
mean(spearmans)
sd(spearmans)

# or use ggplot to display histogram
boot_cor <- data.frame(correlation = c(pearsons, spearmans), 
                       type = c(rep("Pearson's", Nsims), rep("Spearman's", Nsims)))

ggplot(boot_cor, aes(x = correlation, fill=type)) + geom_histogram()
```

**Task** Choose two different continuous variables and bootstrap the Pearson's correlation as well as the Spearman correlation. Plot the histograms along with the 95% confidence interval (from the quantiles). Also add a vertical line for the correlation in the full sample. Do you conclude the correlation is positive, negative, or that there is no clear evidence of correlation?

``` {r}
bootplot = ggplot(subset(boot_cor, type = "Pearson's"), aes(x = correlation)) + geom_histogram()

bootplot = bootplot + geom_vline(xintercept = quantile(pearsons, c(0.025, 0.975))) 
bootplot = bootplot + geom_vline(xintercept = cor(movies$gross, movies$budget), color = "red")
bootplot
```
