---
title: "Summarizing Data"
author: "Data Science 101 Team"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---



# Part I: Robustness of summaries

## Question 1: How many numbers do you need to change in order to change the mean?

Let's start by randomly generating 11 values from the standard normal curve, then compute the
mean of these 11 values:


```{r}
set.seed(1)
sample1 <- rnorm(11, 0, 1) # 11 random data points, mean of the data is 0, standard deviation of distribution is 1
sample1 
```
How many numbers among the 11 values do we need to change so that the new sample mean becomes 2?

# 1 value so the sum of all 11 values becomes 22

```{r}
sample1_mean <- mean(sample1)
sample1_mean
```


## Question 2: How many numbers do we need to change in order to change the median?

We will look at the same data.
```{r}
sample1_median <- median(sample1)
sample1_median
```

- How is this sample_median calculated?
- If we change the largest value to be 10, does the median change? If we change the smallest value to be -10, does the median change?

no!!!!!!!!

- In general, can we get an arbitrary value of median by changing just one value in this sequence? 
no!!!

- How many numbers do we need to change/add if we want a specific median? For example, please answer this question for the two scenarios: new_median = 0.5757, new_median = 1.511.

 1 number and make sure values on one side are greater and smaller on the other side. 


## Question 3: Which one is less sensitive to "extreme" data points, the mean or the median?

# Part II: Loss functions

A loss function measures the distance between a reference point (usually a "center" statistic like mean or median denoted *z* in the lecture slides) and each observation. Another name for the city block distance/absolute error loss is the *L1* loss. Another name for the squared error loss/squared Euclidean distance is the *L2* loss. If **x** is a vector of *N* observations, the loss $L(\mathbf{x}, z)$ is also length *N*. Typically, we are interested in finding the center that minimizes average loss/average distance from the center. The average distance from the center is called *mean absolute error* (MAE) if we use city block distance and average distance from the centeris called *mean squared error* (MSE) if we use squared Euclidean distance.
$$ 

MAE(\mathbf{x}, z) = \sum_{i=1}^{n} |x_i - z|, MSE(\mathbf{x}, x) = \sum_{i=1}^n (x_i - z)^2 \\

mean(\mathbf{x}) = \arg \min_{z} MSE(\mathbf{x}, z), \\ median(\mathbf{x}) = \arg \min_{z} MSE(\mathbf{x}, z)
$$
```{r}
L1_loss <- abs(sample1 - sample1_median)
# mean absolute error 
MAE <- mean(L1_loss)
MAE
```

```{r}
L2_loss <- (sample1 - sample1_mean)^2
MSE <- mean(L2_loss)
MSE
```

If we need to compute the *MAE* repeatedly, then it is useful to define a function that computes it:

```{r}
MAE <- function(sample, center) {
  
  L1_loss <- abs(sample - center)
  mean_L1_loss <- mean(L1_loss)
  
  return(mean_L1_loss)
  
}
MAE(sample1, 1)

```



Now check that we get the same answer as before by calling this function using the data and the median
as arguments:

```{r}
MAE(sample1, sample1_median)
```


**Task**: In lecture, we claimed that the median minimizes *L1* (absolute error) loss and the mean minimizes *L2* (squared error) loss. Let's see if this is true. Generate a vector that represents
a grid from 0.3 to 0.4 with a spacing of 0.001. Evaluate the *MAE* on this grid, i.e. each grid point
serves as a center in turn. Plot the *MAE* as a function of these center values.
 
```{r}
centers <- seq(0.3, 0.4, 0.001)

mae <- c()

for(i in 1:length(centers)) {
  mae[i] <- MAE(sample1, centers[i])
}

```


```{r}
df <- data.frame(x = centers, y = mae)
ggplot(df, aes(x, y )) + geom_point()

plot(centers, mae)
```

Find the minimum of these *MAE*s. Then find the element of centers that attains this minimum:

```{r}
min(mae)
```

The function which.min(x) gives the index of the vector x at which the minimum of x is attained.

TO find the index of the sequence where mae is minimized, call which.min().

```{r}
which.min(mae)
```



```{r}
centers[which.min(mae)]
```

